---
title: "F1 2017 Australia, Qualifying"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
library(knitr)
options(knitr.kable.NA = '')
opts_chunk$set(fig.path = "images/f1_2017_aus-")
```

```{r}
library(rvest)
library(plyr)
library(ggplot2)

library(f1djR)

qlaps= read.csv("~/Dropbox/various/aus17_qualilaptimes.csv")
cutoff=c(15,10,2)
```

```{r}
##Session classifications page
xx=pageGrabber('http://www.fia.com/events/fia-formula-one-world-championship/season-2017/session-classifications-19')
```

```{r}
library(grid)
library(gridExtra)
promoter= function(g) {grid.arrange(g, nrow=1,
             #top="My title", 
             bottom = textGrob("To learn how to create this chart: f1datajunkie.com",
                               gp = gpar(fontface=3, fontsize=9),
                               hjust=1, x=1))}
```
```{r}
qr=quali_progression_ergast_tx(ergastifyQualiClass(fiaSessionClassQualifyingTidy(fiaTableGrabber(xx,4))))
qrm=quali_progression_ergast_melt(qr)
qrtmp=qr
for (i in c(1,2,3)) {

  qrtmp[[paste0('Q',i)]]=ifelse(is.na(qrtmp[[paste0('Q',i)]]),'',
                                paste0(qrtmp[[paste0('Q',i)]],' (',qrtmp[[paste0('q',i,'pos')]],')'))
}
```

Drivers progressed through the qualifying session as follows:
```{r quali_slopegraph}
core_qualifying_rank_slopegraph(qr,qrm,spacer=0.21,reverse=T)
```

The official session results were recorded as follows:
```{r}
kable(subset(qrtmp,select=c('POS','DRIVER','Q1',	'Q2',	'Q3'	)), format='markdown',row.names = FALSE)
rm(qrtmp)
```


## Session Utilisation

Activity throughout the three qualification sessions took place as follows:

```{r quali_utilisation,fig.width=10}
#qlapsa=rawLap_augment_laptimes(qlaps)
qlapsb=rawLap_augment_quali(qlaps)
qlapsb=qsessionOverride(qlapsb,0,1500,2800)
qlapsb=quali_purplePatch(qlapsb)
promoter(augmented_session_utilisation_chart(qlapsb,2,session="F1 AUS 2017, Qualifying"))
```

Q3 was interrupted by a red flag event.

## Q1 Cut-off Time Evolution

The cut-off time in Q1 evolved as follows:

```{r quali1_cutoff}
promoter(quali_session_times_plots(qlapsb[qlapsb['qsession']==1,],1,FALSE,FALSE,TRUE,cutoff)+theme_bw())
```

## Q2 Cut-off Time Evolution

The cut-off time in Q2 evolved as follows:

```{r quali2_cutoff}
promoter(quali_session_times_plots(qlapsb[qlapsb['qsession']==2,],2,FALSE,FALSE,TRUE,cutoff)+theme_bw())
```

## Q3 Cut-off Time Evolution

The cut-off time in Q3 evolved as follows:

```{r quali3_cutoff}
promoter(quali_session_times_plots(qlapsb[qlapsb['qsession']>2,],3,FALSE,FALSE,TRUE,cutoff)+theme_bw())
```

---
 To learn how to wrangle F1 results and timing data to create charts like these, see [f1datajunkie.com](https://f1datajunkie.com) or [buy the *Wrangling F1 Data With R* book](https://leanpub.com/wranglingf1datawithr).
