# Career Trajectory


*Based on the **Career Trajectories** chapter of **Analyzing Baseball Data with R**.*

```{r}
require(RSQLite)

con_ergastdb = dbConnect(drv='SQLite', dbname='./ergastdb13.sqlite')

#tbs=dbGetQuery(con_ergastdb, 'SELECT name FROM sqlite_master WHERE type = "table"')
```


Use crude year count - better to calculate age at the start of the year? Or mid-season?

```{r}
drivers2013=dbGetQuery(con_ergastdb, 'SELECT * FROM drivers WHERE driverId IN (SELECT DISTINCT ds.driverId from driverStandings ds JOIN races r WHERE r.year=2013 AND r.raceId=ds.raceId )')


```

Find the standings at the end of a particular year - 

```{r}
drivers2013=dbGetQuery(con_ergastdb, 'SELECT d.driverRef, d.code, ds.position, d.dob FROM driverStandings ds JOIN drivers d WHERE ds.raceID IN (SELECT raceId FROM races WHERE year=2013 and round IN (SELECT MAX(round) FROM races WHERE year=2013)) AND ds.driverId=d.driverId')

drivers2013$dob=as.Date(drivers2013$dob)
drivers2013$birthyear=format(drivers2013$dob, "%Y")

```


```{r message=T,echo=T}
source('~/Dropbox/wranglingf1datawithr/src/ergastR-core.R')
```

```{r}
alonso = driverCareerStandings.df('alonso')

getYearFromDate=function(date){
  as.numeric(format(as.Date(date), "%Y")) 
}

driverData.list=function(driverRef){
  dURL=paste(API_PATH,'drivers/',driverRef,'.json',sep='')
  drj=getJSONbyURL(dURL)
  dd=drj$MRData$DriverTable$Drivers[[1]]
  list(
    dateOfBirth=as.Date(dd$dateOfBirth),
    driverId=dd$driverId,
    nationality=dd$nationality,
    yearOfBirth=getYearFromDate(as.Date(dd$dateOfBirth))
  )
}
alonsodata=driverData.list('alonso')
drivercareer.aug=function(driverRef){
  drivercareer=driverCareerStandings.df(driverRef)
  driverdata=driverData.list(driverRef)
  drivercareer$age=drivercareer$year-driverdata$yearOfBirth
  drivercareer
}

careerplot=function(driverRef){
  drivercareer=drivercareer.aug(driverRef)
  ggplot(drivercareer)+geom_point(aes(x=age,y=pos))
}

#baseball book
fit.model= function(d){
  fit = lm(pos ~I(age) +I((age)^2),data=d)
  b=coef(fit)
  age.max=30-b[2]/b[3]/2
  maxx=b[1]-b[2]^2/b[3]/4
  list(fit=fit,age.max=age.max,maxx=maxx)
}

fit.model2= function(d){
  fit = lm(pos ~driverRef+I(age) +I((age)^2),data=d)
  b=coef(fit)
  age.max=30-b[2]/b[3]/2
  maxx=b[1]-b[2]^2/b[3]/4
  list(fit=fit,age.max=age.max,maxx=maxx)
}

fit.model3= function(d){
  fit = lm(pos ~driverRef+car+I(age) +I((age)^2),data=d)
  b=coef(fit)
  age.max=30-b[2]/b[3]/2
  maxx=b[1]-b[2]^2/b[3]/4
  list(fit=fit,age.max=age.max,maxx=maxx)
}

al=drivercareer.aug('alonso')
ff=fit.model(al)
plot(al$age,al$pos)
lines(al$age,predict(ff$fit,age=al$age))


but=drivercareer.aug('button')
web=drivercareer.aug('webber')
al=drivercareer.aug('alonso')
aa=rbind(but,web,al)

ff2=fit.model2(aa)
plot(aa$age,aa$pos)
#?how do we plot just the age and age^ parameters?
lines(aa$age,predict(ff2$fit,age=aa$age))

albu2=rbind(al2,bu2)

ggplot(aa,aes(x=age,y=pos))+stat_smooth(method = "lm", formula = y ~ I(x-28) +I((x-28)^2)) +geom_point()

ggplot(aa,aes(x=age,y=pos,col=driverRef))+stat_smooth(method = "lm", formula = y ~ I(x-28) +I((x-28)^2)) +geom_point()

ggplot(aa,aes(x=age,y=pos,col=driverRef))+stat_smooth(method = "lm", formula = y ~ I(x-28) +I((x-28)^2)) +geom_point()+facet_wrap(~car)

ggplot(aa,aes(x=year,y=pos,col=car))+stat_smooth(method = "lm", formula = y ~ I(x-28) +I((x-28)^2)) +geom_point()

#need a team continuity column
require(zoo)
#age in years and months between race and dob
as.yearmon(as.Date('2010-11-14'))-as.yearmon(as.Date('2009-10-14'))
```


##BBC analysis
http://www.bbc.co.uk/sport/0/formula1/26464195

#F1 grandprix
http://f1-grandprix.com/?page_id=3256


http://www.pinnaclesports.com/online-betting-articles/06-2013/formula-one-driver-pole-conversion-rates.aspx


The age-productivity gradient: evidence from a sample of F1 drivers
http://www1.unive.it/media/allegato/DIP/Economia/Working_papers/Working_papers_2009/WP_DSE_castellucci_pica_padula_16_09.pdf

??how about doing all races and age as 


Get drivers in 2013

Get last races of year

Get standings of drivers in 2013 in all last races of year

Get all their standins in last 



FInd last races over all years 

dbGetQuery(con_ergastdb, 'CREATE TEMPORARY VIEW lastRounds AS SELECT s.year,r.raceId,r.round,r.name, s.maxRound FROM (SELECT year,MAX(round) maxRound FROM races GROUP BY year) s JOIN races r WHERE r.round=s.maxRound AND r.year=s.year')

dbGetQuery(con_ergastdb, 'SELECT * from lastRounds LIMIT 10')

dbGetQuery(con_ergastdb, 'CREATE TEMPORARY VIEW drivers2013 AS SELECT * FROM drivers WHERE driverId IN (SELECT DISTINCT ds.driverId from driverStandings ds JOIN races r WHERE r.year=2013 AND r.raceId=ds.raceId )')

dbGetQuery(con_ergastdb, 'CREATE TEMPORARY VIEW constructors2013 AS SELECT * FROM constructors WHERE constructorId IN (SELECT DISTINCT cs.constructorId from constructorStandings cs JOIN races r WHERE r.year=2013 AND r.raceId=cs.raceId )')

dbGetQuery(con_ergastdb, 'SELECT * from drivers2013')
dbGetQuery(con_ergastdb, 'SELECT * from constructors2013')

tmp=dbGetQuery(con_ergastdb, 'SELECT * from driverStandings ds JOIN races r JOIN drivers d WHERE ds.raceId in (SELECT raceId FROM lastRounds) AND ds.driverId IN (SELECT driverId FROM drivers2013) AND r.raceId=ds.raceId AND d.driverId=ds.driverId')

tmp2=dbGetQuery(con_ergastdb, 'SELECT * from constructorStandings cs JOIN races r JOIN constructors c WHERE cs.raceId in (SELECT raceId FROM lastRounds) AND cs.constructorId IN (SELECT constructorId FROM constructors2013) AND r.raceId=cs.raceId AND c.constructorId=cs.constructorId')

drivers with at least 10 years history

dbGetQuery(con_ergastdb, 'SELECT d.driverId, d.driverRef, COUNT(*) years FROM driverStandings ds JOIN drivers d WHERE raceId IN (SELECT raceId FROM lastRounds) AND d.driverId=ds.driverId GROUP BY ds.driverId HAVING years>=10 ORDER BY years DESC')

Can also do chamionships by constructor?

dbGetQuery(con_ergastdb, 'SELECT c.constructorId, c.constructorRef, COUNT(*) years FROM constructorStandings cs JOIN constructors c WHERE raceId IN (SELECT raceId FROM lastRounds) AND c.constructorId=cs.constructorId GROUP BY cs.constructorId HAVING years>=10 ORDER BY years DESC')

??any dbpedia data for teams previously known as?

