```{r}
opts_chunk$set(fig.path = "images/", message=FALSE)
```
# Chapter - Practice Sessions

*In which we learn how to look at charts and go back to basics in terms of how to read them, as well as getting started with constructing some charts of our own.*

Whilst the focus of a race weekend are rightly the Saturday qualifying session and the Sunday race, the race weekend currently begins with P1 and P2, the two, ninety minute Friday practice sessions (Thursday at Monaco), and the hour long third practice (P3) session before qualifying, as specified in article 32 of the [FIA Formula One Sporting Regulations](http://www.fia.com/sport/regulations?f%5B0%5D=field_regulation_category%3A82). The practice sessions provide the teams with the essential track time they need to set up the cars, and in doing so provide a source of data that may help us identify which cars are likely to be competitive on any particular race weekend. In particular, second practice provides the teams with an opportunity to try some long run race simulations. (As the intelligentf1 blog shows, analysis of the laptimes from these second practice long runs can often provide a good insight in the underlying race pace and performance of each car.) Third practice, occurring as it does a couple of hours before qualifying (article 32 of the 2014 regulations set third practice (P3) time between 11.00 and 12.00, with qualifying scheduled between 14.00 and 15.00) provides teams with a final chance to set the car up for qualifying.

The ergast database does not contain any data relating to the practice sessions, but we can get a certain amount of information from the formulaOne.com results pages, as well as more detailed timing information from the FIA media centre. Prior to the 2014 season, the free (though login require) live timing application on the Formula One website also used to stream live sector times; using third party applications, this data could be scraped and archived from the live timing feed. However, the 2014 season has seen functionality on the free live timing app reduced, inclusing the withdrawl of sector times. At the time of writing, third party developers have not publicly released reverse engineered applications capable of streaming (and logging) data from the paid for official Formula One live timing appliction.


//sidebar
The intelligentf1.com website provides a great example of how we can make use of lap time data from the race simulations that are typically carried out by each team in second practice. On that site, James Beck uses the laptime data to calibrate a model that explores the relative competitiveness of each driver and give a feel for the relative tyre degradation rates experienced by each team.

*Throughout this chapter, we will focus solely on data scraped from the Formula One website. Later editions of this book may also include analysis of data obtained from the official FIA timing sheets.*

To start with, we need to establish a connection to the database containing the scraped data.

```{r message=FALSE}
library("RSQLite")
f1comDatabaseConnection=function(
                          drv="SQLite",
                          db="~/Dropbox/wranglingf1datawithr/src/scraperwiki.sqlite"
                        ){ dbConnect(drv=drv, dbname=db) }
f1=f1comDatabaseConnection()
```

The structure of each of the practice tables are the same. The tables containing results from a practice session take the following form:

```{r results='asis'}
kable(dbGetQuery(f1, ("SELECT * FROM p1Results LIMIT 5")))
```

The practice session results include the name of each driver; their classification within that session; their team; the number of laps they completed; their best laptime as a natural time (using the format *minutes:seconds.milliseconds*) and as a time in seconds and milliseconds; and the natural gap (*natgap*)/*gap* (the *natgap* as seconds/milliseconds) to the best time in the session.

Q> What can we do with the practice session results data?
Q> Spend a few minutes sketching out what charts or analyses you think you might be able generate from this data. For example, which data columns can usefully be plotted against each other, and what would such visualisations show? How might data be grouped, either as the basis of point colouring or for within group comparisons or analysis? How might you combine - and visualise - data from all the practice sessions in a race weeked? What *derived data columns* can you generate from the data?

To start with, we will consider charts that correspond to data from a single session in a single race. Then we will have a look at what additional views we may get from the data by comparing and/or aggregating data from two or three practice sessions from the same race weekend.

Trivially, in the case of a single practice session from a single weekend, we might begin with:

* the best laptime by driver versus driver classification in a session
* the number of laps completed by each driver
* a single dimensional chart show the position of each driver's laptime along a timeline.

We can quickly sketch these as follows:

```{r}
#Get some data - P1 from Malaysia in 2012
dd=dbGetQuery(f1, ("SELECT * FROM p1Results WHERE year=2012 AND race='MALAYSIA'"))

#Check the column types
#str(dd)
#It may be the case that not all the columns that should be integers actually are...
#If necessary, use this temporary hack till I revise the scraper that fills the database
#dd$pos=as.integer(dd$pos)
#dd$laps=as.integer(dd$laps)

library(ggplot2)
ggplot(dd)+geom_point(aes(x=pos,y=time,col=team)) 
```

This chart shows how the times differ by rank. We use the colour attribute to help us spot whether there appears to be groupng based on team.

The simplest view over the lap count is a bar chart. The biggest question this initial sketch raises for me is how we should order the horizontal x-axis, which identifies each driver.

```{r}
ggplot(dd)+geom_bar(aes(x=driverNum,y=laps),stat='identity')
```

Q> ###What, if anything, is wrong with these charts?
Q> In both charts, the values on the x-axes are arranged in "alphabetical" rather than numerical order. In the case of driver number, this may make sense (the driver number is essentially an arbitrary, unique categorical label or identifier), but it is a nonsense as far as the position ordering goes.
Q> In the laptime chart, the intention behind the x-axis ordering was to order times by position, which does not give us any imformation within the chart about which driver the time was associated with. (We do, however, get information about the team from the point colour.)
Q> It is also worth noting in the laptime chart that the x-axis describes a discrete scale, with markers equally spaced horizontally, whereas the y-axis is a continuous scale, and the vertical spacing between points is related to the actual difference in y-value represented by each point. 
Q> In the lap count chart, the y-axis values are also "alphabetically"" ordered. This really is a nonsense and shows how important it is to check your axis tick mark labels. To make this chart in any way meaningful, we need to ensure that the lap count is treated as a numerical quantity.

Insofar as as chart represents a particular quantity associated with a particular car or driver, what might the most appropriate ranking be? The first decision to take is whether the x-axis values are ordered based on the x-axis values *or the y-axis values*. Here are some possibilities:

* for the laptime chart, the most sensible classification is probably in rank order of position, which is also the rank order of the fastest laptime (the y-axis value);
* for the lap count chart, we could rank by driver number. This arguably made more sense using the numbering scheme that existed prior to 2014, when personal numbers were introduced, because drivers were essentially ranked based on performance of the team in the Constructors' Championship from the previous season and the nominal first driver/second driver ranking within each team). With personal driver numbers the numerical ordering is not really meaningful - the personal driver numbers are qualitiative, categorical labels rather than numbers *per se*. A more useful ordering would probably be to order the x-axis values so that the y-values are in height order (that is, by increasing or decreasing lap count), relabelling the x-axis *driverNum* with the corresponding driver's name or identifier.
* we might arrange the bars into groups for each team so we see at a glance how much track time each team had. But if we do that, how do we order the groups, and the bars within each team? Should we dodge them or stack them?

Whenever you have a categorical axis, such as driver identifiers in the case of the lap count chart, it's worth remembering that the ordering is essentially arbitrary, although some orderings may be more meaningful than others in terms of the semantic relationships between the different categorical values. (For example, team members group well; previous world champions vs rookies vs others is another possible grouping; and so on.) Different ordering choices may also help you see different patterns or structures within the data.

If we inspect the structure of the data, we see that the `pos` column is indeed identified as a character type, as are the `laps` and `driverNum` columns:

```{r}
str(dbGetQuery(f1, ("SELECT * FROM p1Results LIMIT 5"))[c('pos','laps','driverNum')])
```

If we recast the postion to an integer, the laptime chart can be represented more intuitively (we can also take the opportunity to tidy up that chart labels):

```{r}
dd$pos=as.integer(dd$pos)

g=ggplot(dd)+geom_point(aes(x=pos,y=time,col=team)) 
g=g+xlab('Position')+ylab('Time (s)')+ggtitle('F1 Malaysia, 2012: FP1 Laptimes')
# We can set the title of the legend, or remove it, as in this case.
# Note the default ordering in the legend is alphabetical and a default
# colour palette is used.
g=g+ guides(fill=guide_legend(title=NULL))
g
```

By eye, we can identfy several different groupings in the data. Let's add some guiding lines to the chart to separate out the groups: the first two cars are clear of cars in positions 3 to 10; cars 11 to 18, or perhaps 11 to 20 are in a group of their own, and then the four backmarkers are way off the pace.

```{r}
g2=g+geom_hline(yintercept=98.65,col='grey')
g2=g2+geom_hline(yintercept=99.6,col='grey')
g2=g2+geom_hline(yintercept=100.7,col='grey')
g2=g2+geom_hline(yintercept=101.3,col='grey')
g2
```

One thing we might want to think about it how to identify clusters more reliably/alogorithmically.

Another way of viewing the same data is to look at the lap times in terms of the percentage they are over the fastest laptime. The 107% value is meaningful in the sense of F1 regulations, because if a car does not record a time within 107% of the first placed car in qualification, it does not automatically qualify for the actual race.

```{r}
dd$percent=100*dd$time/min(dd$time)
ggplot(dd) + geom_point(aes(x=pos,y=percent),stat='identity') + geom_hline(yintercept=107,col='grey')+ylim(100,110)
```

To get a clearer view of the time differences between cars, we can look at the difference in time between each car and the car classified one position ahead of it by generating a `delta` column between their lap or gap to leader times. To do this, sort the data by position and then use the `diff()` function to find the lagged difference between consecutive times in the time/position ordered dataframe.

```{r results='asis'}
require(plyr)
ddx=dd
ddx=arrange(ddx,pos)
ddx$delta=c(0,diff(ddx$natGap))

kable(head(ddx,n=3))
```

We can now plot these values as a bar chart, adding a line that helps us identifie deltas of more than a quarter of a second between cars.

```{r}
ggplot(ddx) + geom_bar(aes(x=pos,y=delta),stat='identity') + geom_hline(yintercept=0.25,col='grey')
```

Note the way that ggplot constructs the chart by adding each consecutive element as a higher layer in the chart. If we want the horizontal line *below* the bars, we can simply reorder the way in which we add the layers to the chart. Let's also colour the bars differently depending on whether the delta to the next car is within the 0.25s boundary.

```{r}
ggplot(ddx) + geom_hline(yintercept=0.25,col='grey') + geom_bar(aes(x=pos,y=delta,col=factor(delta<0.25)),stat='identity')
```

Returning to the laps count, how might we more correctly render that? Recall, the `laps` count wasn't being correclty rendered as a number, and we also needed to find a way to sort the x-axis values on the basis of the corresponding y-axis value. In the following example, we sort in terms of decreasing lap count.

```{r}
dd$laps = as.integer(dd$laps)
g = ggplot(dd)+geom_bar(aes(x=reorder(driverName,-laps),y=laps),stat='identity') + theme(axis.text.x = element_text(angle = 45, hjust = 1))
g
```

We can also rotate the chart to provide a horizontal bar chart view, which is arguably easier to read in this case.

```{r}
g + coord_flip()
```

In terms of looking for meaning in this chart, we might look to see how many laps the top 5 cars by laptime completed. We can highlight the drivers classified in the top 5, adopt a clean, minimal theme and then suppress the legend.

```{r}
#In part via http://stackoverflow.com/a/12910865/454773
dd[["top5"]] = ifelse(dd$pos <= 5, "top5", "notTop5")
ggplot(dd) + geom_bar(aes(x=reorder(driverName,-laps),y=laps,fill=top5),stat='identity') + coord_flip()+ylab(NULL)+xlab('Laps') + scale_fill_manual(values = c("top5" = "darkblue", "notTop5" = "lightgrey")) + theme_classic() + theme(legend.position="none")
```

If we want to compare the number of laps by team, we can group the bars on that basis. Prior to 2014, we could use driver number to identify which driver was the "first" driver in a team and which driver was the second ranked driver and then order drivers within teams on that basis. One way to do this for the current season might be to nominate the driver who was classified highest in the previous year's Drivers' Championship as the "first driver" within a team (with rookies being ranked lowest and a stategy for numbering drivers in teams with two rookies). In the chart below, it is ambiguous which driver corresponds to which bar within each team. For visual clarity, it might make sense to order the bars within each group in ascending order:

```{r}
#Order the driverName used for grouping within each team by number of laps
dd$driverName=reorder(dd$driverName,dd$laps)

ggplot(dd)+geom_bar(aes(x=team,y=laps,group=factor(driverName)),stat='identity',fill='grey',position='dodge')+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+xlab(NULL)+ylab('Laps')
```

To view the total number of laps by team, we can stack the bars, perhaps ordering the chart by the total number of laps completed by the team.

```{r}
ggplot(dd)+geom_bar(aes(x=reorder(team,laps),y=laps),stat='identity',fill='grey',position='stack')+ theme(axis.text.x = element_text(angle = 45, hjust = 1))+xlab(NULL)+ylab('Laps')
```

As you might imagine, if we had a single dataframe containing the number of laps completed by drivers from each team from across several practice sessions, we could use this sort of chart to display the total number of laps of running achieved by each team throughout practice.

Let's flip back again to the laptimes chart: how else might we try to identify not only the gap between drivers, but also their positions, by driver name? One way is to use a horizontal x-axis that identifies the laptime, and then place a marker for each driver, identiying them by name. This essentially reduces the data to a single dimension data set. In the chart below, we space driver names and markers according to lap time. Note how we rotate the driver names to make the chart more readable as well explicitly indicating the relative laptimes.

```{r}
ggplot(dd)+geom_point(aes(x=percent,y="Time"))+geom_text(aes(x=percent,y="Name",label=driverName),angle=90,size=3)+ theme_classic() 
```

Whilst there are some problems with this chart (for example, some occlusion of overlapping names) the linear dot plot once again highlights the 

X> ### Comparing Team Performances
X> How would you modify the above chart to display the team names rather than driver names, organising the names using a vertical rather than horzontal profile? Can you think of a way of making the names more readable?

We can use team names rather than driver names by setting the text label to the `team` column. To rotate the chart, we can use `coord_flip()`, and also remove the label rotation within the `geom_text()` element.

One way of making the names more readable is to split the data into separate columns, for example putting cars placed in odd numbered positions in one column and those placed in even numbered comlumns in a second column.

```{r}
#Basic dot plot
g=ggplot(dd) + geom_point(aes(x=time,y="Time"))
#Split the drivers into two groups - odd position number and even position number
#Use each group as a separate y-axis categorical value
g=g+geom_text(data=subset(dd,subset=(pos %% 2!=0)),aes(x=time,y="1,3,5,...",label=team),size=3)
g=g+geom_text(data=subset(dd,subset=(pos %% 2==0)),aes(x=time,y="2,4,6,...",label=team),size=3)
#Tweak the theme
g=g+theme_classic() + ylab(NULL)
#Add in some intercept lines using the values we used before
g=g+geom_vline(xintercept=98.65,col='grey')+geom_vline(xintercept=99.6,col='grey')+geom_vline(xintercept=100.7,col='grey')+geom_vline(xintercept=101.3,col='grey')
#Flip the co-ordinates
g=g+coord_flip() 
g
```

If team names are close to each other on a horizontal line or vertically, then those cars are close in terms of time, even if not directly next to each other in terms of position. Once again, the grouping elements are also emphasised by visual proximity, something we can additionally reinforce using the grey splitter lines. Although the grouping is to some extent arbitrary, it may be useful as a storytelling or journalistic device.

Something else that jumps out from this particular example is that the team names are actually a combination of the manufacturer and engine supplier. (Where the name is not hyphenated the manufacturer *is* the engine supplier, as in the case of Ferrari and Mercedes). This suggests we can generate an additional data column - *engine*.

```{r,results='asis'}
#For each row, split the team name on a '-' and take the last item in the resulting list
dd=ddply(dd, .(driverName),mutate,engine = tail(strsplit(team,'-')[[1]],n=1) )
kable(head(dd,n=3))
```

We can quickly count how many teams run each engine. First, identify the unique engine/team combinations:

```{r}
engineTeams=unique(dd[, c('engine','team')])
engineTeams
```

Then count the number of occurrences of each engine:

```{r}
#Use the plyr count() function
count(engineTeams,'engine')
```

We can also generate a summary report of the total, mean (average) and median number of laps completed by each car running each engine type:

```{r}
ddply(dd,.(engine),summarise,totLaps=sum(laps),meanLaps=mean(laps),medianLaps=median(laps))
```

We can also make use of the engine type to colour our laptime ordering chart.

```{r}
#Basic dot plot
g=ggplot(dd) + geom_point(aes(x=time,y="Time"))
#Split the drivers into two groups - odd position number and even position number
#Use each group as a separate y-axis categorical value
#Add in an additional colour aesthetic tied to engine type
g=g+geom_text(data=subset(dd,subset=(pos %% 2!=0)),aes(x=time,y="1,3,5,...",col=engine, label=team),size=3)
g=g+geom_text(data=subset(dd,subset=(pos %% 2==0)),aes(x=time,y="2,4,6,...",col=engine, label=team),size=3)
#Tweak the theme
g=g+theme_classic() + ylab(NULL)
#Add in some intercept lines using the values we used before
g=g+geom_vline(xintercept=98.65,col='grey')+geom_vline(xintercept=99.6,col='grey')+geom_vline(xintercept=100.7,col='grey')+geom_vline(xintercept=101.3,col='grey')
#Flip the co-ordinates
g=g+coord_flip() 
g
```

Although only an informal sort of analysis, our eye suggests that in this case Mercedes and Renault powered cars were competitive with each other, Ferrari powered cars were slightly behind, and the Cosworth powered teams were in a league of their own, though in the wrong direction!

X> ### Finding Gaps
X> One of the things the above chart requires is the explicit placement of the lines that separate identifiable-to-the-eye groupings. It would be nice if we could come up with some sort of heuristic to automatically place these. For example, we might identify gaps between consecutive positions above a particular threshold (using something like `ddx$deltadelta = c(0, diff(ddx$delta))` perhaps?) and place a line mid-way through the gap. To prevent grouping singletons, we would also need to count how many cars are in each grouping and perhaps *not* draw a line if there is less than particular number of cars in a group, or the difference between consecutive gaps falls below some threshold value. 

Hopefully these exampltes show how even the simplest data tables can be mined to produce a wide variety of differnt summary reports in both tabular and graphical form.

## Sector Times

The data in the practice session sectors tables contains the sector number, the driver number, their position in the session classification, their name, and their best recorded sector time in that session.

```{r results='asis'}
kable(dbGetQuery(f1, ("SELECT * FROM p1Sectors LIMIT 5")))
```

As Ricardo Divila writes in *Going with your gut*, his RaceCar Engineering column from June 2014 (p5), *"[s]ectors that repeat are ones without problems for drivers - the ones that vary are the difficult ones and show where car setup problems lie."* Unfortunately, the FIA don't publish the full history of sector times, so this form of insight is not available to us. (Prior to 2014, the free FIA timing app had been a source of scrapeable sector times via third party timing apps that have reverse engineered the live timing data feed. As of the start of the 2014 season, this data seems no longer to be streamed through the free app.)

We work with what we do have, however. Inspecting the structure of the data, we see that not all the columns are typed as naturally as we might like:

```{r}
str(dbGetQuery(f1, ("SELECT * FROM p1Sectors LIMIT 5")))
```

To correct this we need to cast the column types explicitly:

```{r}
p1sectors=dbGetQuery(f1, ("SELECT * FROM p1Sectors"))
p1sectors$sectortime=as.double(p1sectors$sectortime)
p1sectors$pos=as.integer(p1sectors$year)
p1sectors$pos=as.integer(p1sectors$pos)
str(p1sectors)
```

Using the individual sector times, we can calculate the *ultimate lap* for each driver as the sum of their best sector times.

To do this we need to generate the sum of the sector times recorded for each driver in each race of each year.

In pseudo-code, we might imagine a recipe for achieving this sort of operation taking the form:

```for each year:
  for each race:
    for each driver:
      calculate the sum of the driver's sector times```


An alternative way to approach the same calculation is to adopt a *split-apply-combine* strategy, as described in ???. Using the `plyr` library, we can call on a particular function, `ddply`, that allows us to split a data frame into groups based on the values of one or more columns, and then perform a summarising operation across the members of each grouping.

In this case, we need to split the data into groups corresponding to the data rows associated with each particular driver in each particular race of each year. This should result in three rows for each group, one row for each of the three sectors. The summarising operation we then need to perform is to summarise the data in each group by calculating the sum of the sector times within that grouping.

```{r results='asis'}
library(plyr)
ultimate=function(d) {
  ddply(d, c("year","race","driverName"),summarise,ultimate=sum(sectortime,na.rm=T))
}
ult=ultimate(p1sectors)
kable(head(ult))
```

We can then compare the ultimate laptimes for each driver in a session to the best laptime they recorded in the session. (Note that it may not be possible to drive such an ultimate lap - the best sector time achieved for a particular sector may require taking a line that jeapardises another.)

Let's just check the data we can pull in from the session results tables:

```{r}
p1results=dbGetQuery(f1, ("SELECT * FROM p1Results"))
p1results$laps=as.integer(p1results$laps)
str(p1results)
```

One way of comparing the ultimate lap time with actual laptime for each driver in a particular race is to plot the result on to a two dimensional scatterplot. The easiest way to plot this is from a dataframe that contains the data corresponding to the ultimate lap time in one column and the actual laptime in another.

At the moment, we have the data in two separate tables, *ult* and *p1results*. We can merge the data into a single dataframe using the *year*, *race* and *driverName* columns as merge keys:

```{r results='asis'}
p1results_merge=merge(p1results, ult, by=c("year","race","driverName"))
kable(head(p1results_merge,n=5))
```

We can then plot directly from the merged dataset. Let's take a subset of the data, focussing on the 2012 Australian Grand Prix:

```{r results='asis'}
gp_2012_aus_p1_results = subset(p1results_merge,year=='2012' & race=="AUSTRALIA")
kable( gp_2012_aus_p1_results )
```

```{r}
library(ggplot2)
ggplot(gp_2012_aus_p1_results)+geom_point(aes(x=time,y=ultimate))
```

So what's wrong with this chart? Two things immediately come to mind. Firstly, there is an outlier: one of the drivers appears not to have a session time recorded. Unfortunately, which don't know which driver this time occurs for, which leads to the second problem: which point corresponds to which driver? 

One way of addressing the outlier problem is to filter out drivers for whom no time is recorded in the session (if they do record a time, they will also necessarily have separate sector times, and hence an ultimate laptime, recorded). We can also check that the time is not recorded as absent, that is, as `NA`.

```{r}
gp_2012_aus_p1_results = subset(gp_2012_aus_p1_results, time>0 & !is.na(time) )
```

To address the other issue, that of not knowing which driver each point refers to, we can instead use a text plot. This requires using an extra aesthetic parameter, `label`, that identifies which column's values should be displayed as the text label for each plotted marker.

```{r}
ggplot(gp_2012_aus_p1_results)+geom_text(aes(x=time,y=ultimate,label=driverName))
```

This chart suggests that the ultimate times broadly follow the session times, which makes sense. However, it's virtually impossible to tell whether a driver's session time matched their ultimate time, or whether it was some way away from it. Several factors contribute to this lack of clarity:

* we can't tell what is being used as the registration point for each label - that is, which part of the label marks the `(time, ultimate)` co-ordinates.
* the length of the labels covers a wide range. If the registration point is the mid-point of the label, where is that exactly?
* the font size used for the labels is quite large, meaning that labels obscure each other;
* the name labels have overflowed the plotting area;
* some of the labels appear to fall outside the area displayed by the chart, making them difficult to read;
* it's hard to tell where the line corresponding to equal ultimate and session laptimes lies. The grid is probably too coarse grained to be able to take accurate measurements for each marker, even if we could tell where the registration point is.

Let's work through the problems one at a time. ggplot supports layering in plots, with the layer order determined by the order in which layers are added to the plot. By default, the first layer is the lowest layer, the last layer the highest. We can mark the registration point using a `geom_point()` layer.

Let's add the point *underneath* the corresponding label. This means adding the `geom_point()` to a lower level than the `geom_text()` by adding it to the ggplot chart *before* the `geom_text()`.

Whilst we could identify the values to be used as *x* and *y* aesthetics in each layer, we can also declare them in the base plot and allow their values to be inherited by the chart layers

We can also assign the plot to a variable, and build it up a layer at a time, before plotting the final compound chart.

```{r}
g=ggplot(gp_2012_aus_p1_results,aes(x=time,y=ultimate))

#Add the points layer
g=g+geom_point()

#Add the text layer on top of the chart
g=g+geom_text(aes(label=driverName))

#Plot the chart
g
```

To find the line on which the best laptime *equals* the ultimate laptime, we need to draw a line x=y. In ggplot, we can use a `geom_abline()` to draw such a line. BY default, the colour of the line is black, but that will dominate the chart somewhat so I am going to give it a grey colour.

```{r}
g=g+geom_abline(col='grey')
g
```

It makes sense to think of the best laptime as equalling or *falling short* of the ultimate laptime, but how do we read this chart? If the point lies on the line, the laptime equals the the ultimate time. So how do we read this chart to find how far off the ultimate time a particular laptime was?

Consider this example - laptime = 100, ultimate time =9


It may seem picky, but it's important to take care when reading a chart so that you read it as you intended. This is one of the great challenges of data visualisation - producing charts that have what we might term a "natural reading" that is makes sense 'at a glance' and doesn't catch out the unwary reader.

If we are looking for drivers that have missed their ultimate laptime by a considerable margin, having labels for *each* driver displayed confuses the view. How about if instead we just label the drivers who miss the ultimate laptime by more than a specified amount, such as a two or three tenths or more?


```{r}
g=ggplot(gp_2012_aus_p1_results,aes(x=time,y=ultimate))
g=g+geom_point()

#Add the text layer on top of the chart for drivers that miss their ultimate
#laptime by at least 0.3s
#We can also reduce the size of the text label a little
g=g+geom_text(data=subset(gp_2012_aus_p1_results,(time-ultimate)>=0.3),aes(label=driverName),size=3)
g=g+geom_abline(col='grey')
g
```

That's a little clearer, but we still have the problem that the text labels overflow the edges of the chart somewhat.


??rankings of the session laptime and ultimate laptime

radar chart showing team ranking for each sector
http://stackoverflow.com/questions/9614433/creating-radar-chart-a-k-a-star-plot-spider-plot-using-ggplot2-in-r


We can also generate a range of other reference laptimes, such as:

* the best overall session laptime;
* the best overall session laptime in a team;
* the best ulimate lap in a team;
* the ultimate lap for a team (based on the best recorded sector times across a team);
* the ultimate ultimate lap, based on the best sector times recorded across all drivers.




```{r}

```

## FIA Media Centre Timing Sheets

Additional data is available from the FIA media centre in the form of timing data sheets for the practice sessions. Two documents are produced for each session:

* the session classification, which includes the number of laps completed, the fastest lap time recorded by each driver and the time of day at which it was recorded and the speed (though it is not clear to me what this speed represents, exactly?);

* for each driver, the time of day their first lap of the session was recorded, and the separate lap times thereafter; pit events are also recorded alongside the corresponding lap times. The fastest lap time recorded in the session for each driver is highlighted in a bold font.

In order to make use of this data, we need to scrape the data from the PDF timing sheets and place them into a datastore.

This will be explored in a later edition of this book.

