# An Introduction to RStudio and R dataframes

In this chapter you will get to meet the RStudio interactive development environment and start to explore the R language. Whilst this book is not intended to take on the role of teaching you how to become a professional R programmer, you will hopefully learn enough to be able to read and write simple bits of R code yourself, as well as learning how to use some powerful R libraries such as the `ggplot2` graphics library to produce your own data visualisations.

In the majority of cases, wherever a data table, data visulisation or data analysis appears in this book, it will be preceded by the code required to generate it. This in part is a result of the workflow I have chosen to generate the book. Each chapter is written as a an "R markdown" (*Rmd*) formatted document that combines free text styled using the simple *markdown* language and blocks of R code. The *Rmd* document can then be processed so that each block of R code is executed and the results that it generates, such as a table or chart, included in the manuscript.

If you want to replicate any of the charts or analyses included in this book, you should be able to just copy the appropriate preceding bits of code and run them in your own version of RStudio.

## Getting Started with RStudio

RStudio is a free, open source, cross-platform application that runs on Apple OS/X, Microsoft Windows and Linux operating systems. It can also be run as a service on a virtual or cloud hosted machine and accessed remotely via a web browser.

To get RStudio up and running, you need to:

* **download and install R** from the [R Project for Statistical Computing](http://www.r-project.org/). The [download area](http://cran.r-project.org/mirrors.html) requires you to select a "CRAN mirror". *CRAN* is the *Comprehensive R Archive Network* which hosts the R source code and a wide selection of community developed packages that cover a huge range of statistical analysis and data visualisation techniques. CRAN services are located all round the world, so choose a location that is convenient for you.
* **download and install the desktop edition of RStudio** from [RStudio.org](https://www.rstudio.com/ide/download/). 

*Configuring the server edition of RStudio so that you can run it as a service and access it via a web browser is a rather more involved process and will not be covered in this book. If you would like to run the server, the easiest way is probably to define a virtual machine or make use of an Amazon machine image, such as the [RStudio AMI shared by Louis Aslett](http://www.louisaslett.com/RStudio_AMI/).*

Once you have installed RStudio and got it running, you should be presented with something like this:

## Getting Started with R

As a programming language, R uses a syntax that has many resemblances to other programming languages (as well as a few quirks!). In this section, we will review some of the features of the R language that are particularly useful when it comes to working with data. Where appropriate, we will review particular techniques in more detail in later chapters where the technique plays a key role in wrangling the data so that we can visualise or analyse it. For now, the following whirlwind tour should give you an idea of some of the sorts of the thing that are possible, along with a glimpse of how we might achieve them.

Note that there is no trickery involved - the output is generated by running only and exactly the code shown in this document. Indeed, the output from running each code block is automatically inserted into the original manuscript of this book following execution of the code block.

One of the key ideas of many programming languages is the idea of a *function*. A function can be used to run one or more operations over a data set that is passed to it. We *call* a function by writing its name followed by a pair of brackets - for example *getwd()*. This function *returns* the name of, and path to, the directory we're currently  working in.

```{r echo=TRUE}
getwd()
```

To get some help information about a function, in the Rstudio console enter a ? immediately followed by the function name. For example, we can look up what the complemtary `setwd()` function does by entering `?setwd`:


That is, we can use `setwd()` to set the current working directory.

When working with data, we often want to be able to create a list of values. The function `c()` accepts a list of comma separated values that it will then use to construct a vector or a list:

```{r}
c("Hamilton","Rosberg","Vettel","Ricciardo","Alonso","Raikkonen","Button","Magnussen")
```

The margin numbers in the output identify the index number, or list count number, of the first item in each row of the display. In this case, I can see (wor quickly work out) that there are 8 elements in the list.

As most of the datasets we're going to be working with come in the form of two dimensional data tables (the sort of layout you may be familiar with from looking at a simple spreadsheet), then the *data frame* is the data structure you're likely to spend most of your time working with. 

In many respects, dataframes can be thought of in much the same way as a set of tabular data contained in a worksheet in a spreadsheet application. The data is arranged in rows and columns, each of which can be individually identified. In addition, each cell in the dataframe/worksheet can be uniquely addressed by giving its row and column "co-ordinates".

To explore some of the features of a dataframe, I have produced a sample dataset that contains data from the first three races if the 2014 season, identifying who was on the podium for each race along along with their finishing position.

The data is currently stored in a CSV - comma separated variable/value - file, a simple text based file format for sharing tabular data that you can find here: [winners2014start-csv.csv](https://gist.github.com/psychemedia/11187809#file-winners2014start-csv). You can view the raw data file by clicking on the <> icon in the file title bar.


If you download the raw data to your computer, and take note of the location where you actually saved the file too, you can load the data into RStudio:



You can also use the `read.csv()` function to load the data in yourself:

```{r echo=TRUE}
winners=read.csv('~/Dropbox/wranglingf1datawithr/src/winners2014start.csv')
winners
```

Note that in R the operator used to assign a value or an object (such as the output of the `read.csv()` function), to a *variable*, such as `winners` is traditionally written as <-, as in:

```{r echo=TRUE}
winners <- read.csv('~/Dropbox/wranglingf1datawithr/src/winners2014start.csv')
winners
```

However, I prefer to use the equally valid assignment operator = which is conventionally used in many other programming languages.

If we just enter the name of a variable, such as `winners`, the R interpreter will print out the contents of the variable.

If we want to reference just the values in a single column as a vector of values, we can reference the column as follows:

* `DATAFRAME['COLNAME']`: for example, *winners['driverId']*

```{r}
winners['driverId']
```

We can also get a list of the values contained within a column in one of two ways:

* `DATAFRAM[['COLNAME']]`: for example, *winner[['driverId']]*
* `DATAFRAME$COLNAME`: for example, *winners$driverId*

```{r}
winners[['driverId']]
```

If we wish to inspect the contents of a dataframe in particular in a more convenient way, we can using the viewing area of RStudio. The easiest way to do this is to identify which data object you wish to inspect via the *Environment* tab, then click on it to launch it into a tab in the viewing area:

Alternatively, via the RStudio console, you can execute the `View()` function applied to the required dataframe object directly (for example, `View(winners)`).

We can get different sorts of summary about a data from by using the `str()` and `summary()` commands.

To review the structure of a dataframe, use `str()`:

```{r echo=TRUE}
str(winners)
```

In this case, we see that several columns have been *typed* as having **int**eger value contents, (for example, the *carNum* and *pos* columns), *fastlaptime* has been identified as a **num**eric, and the *driverId*, *constructorId* and *race* columns as *Factors* (that is, categorical variables). Since the adoption of personalised, permanent driver numbers for cars at the start of the 2014 season, the *carNum* is less an integer than a categorical variable. We can covert it to a factor as follows:

```{r}
winners$carNum=factor(winners$carNum)
```

If you rerun the `str()` command you will see that the *carNum* is now a factor with 6 levels:

```{r}
str(winners$carNum)
```

If appropriate, we can also cast columns to integer values (with *as.integer()*), numerical values (floats/reals, *as.numeric()*) and to characters/strings (*as.character()*). Note that when casting something detected as a factor to an integer or numeric, we first need to cast it to a character string. So for example, to cast the *winners$carNum* from a factor back to an integer, we would use the construction *as.integer(as.character(winners$carNum))*.

We can inspect the different values take by a categorical variable using the `levels()` function as applied to a column associated with a dataframe (*dataframe$column*):

```{r echo=TRUE}
levels(winners$race)
```

To get summary statistics back about the contents of a dataframe, which may or may not be *meaningful* summary statistics, use `summary()`:

```{r echo=TRUE}
summary(winners)
```

The `read.csv()` function can also load data in from a URL, rather than a path that specifies local file on your own computer. However, the data file is being served from a secure *https* link, rather than a simpler *http* web address, so we need to do a workaround by loading the data in using a function that is contained in the *RCurl* R package. If your R installation does not have the RCurl package installed, you will need to install it yourself. In RStudio, you can install a package from the *Packages* tab:

We can also define a simple function that will check to see whether a package is installed before we try to load it in; if it's missing, the package installer will be run first:

```{r message=FALSE,warning=FALSE}
#Lines preceded by a # are comments that are not executed as R code

# This simple recipe takes in a list of packages that need to be loaded,
# checks to see whether they are installed, installs any that are missing,
# including their dependencies, and then loads them all in

#The list of packages to be loaded
list.of.packages <- c("RCurl","ggplot2")

#You should be able to simply reuse the following lines of code as is
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages,function(x){library(x,character.only=TRUE)})
```

If you grab the link for the *raw* file (the <> icon in the file title bar), we can now use the following recipe to load the data in directly from the URL.

```{r echo=TRUE}
urlstub='https://gist.githubusercontent.com/psychemedia'

#The actual path may be different - check the actual link from
## https://gist.github.com/psychemedia/11187809#file-winners2014start-csv
urlpath='11187809/raw/38295d24d5e2e35e0b92cb2cf4082a500a691ffd/winners2014start.csv'

#The paste() command concatenates a comma separated list values
#joining them with whatever is specified by the sep='' parameter
#Set sep='' for a seamless join
url=paste(urlstub,urlpath,sep='/')
  
library(RCurl)
data = getURL(url)
winners = read.csv(text = data)
winners
```

If we want to pass something into the function, we put it inside the brackets. For example, the *head()* function can take in a dataframe and then display the first few lines of the dataframe (for example, `head(winners)`). The *head()* function can also accept further arguments that are placed within the brackets, and separated by commas. For example, by default, *head()* previews the first 10 rows of a data frame. We can change this number by means of the *n* parameter. To display just the first three (3) lines of the `winners` dataframe, we would write *head(winners, n=3)*.

```{r }
head(winners, n=3)
```

We can also look at the final few lines of a data frame using a complementary function, *tail()*; once again, the *n* parameter can be used to specify how many lines to display.

For the purposes of this book, I am using a further function - `kable()` -   that tidies up the display of data tables. *(`kable()` can be found in the *knitr* library that is preinstalled in RStudio when working with R markdown.Rmd files.)* If you are trying out the code yourself, omit the `kable()` function from it unless you particularly want to the markdown out format.

```{r results='asis'}
kable( head(winners) ,row.names=F,format="markdown")
```

If you have done any programming before, you will probably be familiar with the notion of variable *types*. For example, a variable may be of an *integer* type (that is, it only and must take on whole number values), a *character* type (often referred to as a *string*) or a ral or floating point number. In the same waym the columns within a data frame can take on a particular type.

We can inspect the structure of the table using the command *str()* which gives us some summary information about the type of each column in a dataframe:

```{r}
str(winners)
```

You will notice that in this dataframe, the columns are made up of *integer* types and *factors*. Factors can be thought of as *categorical variables*, where the values come from an enumerated list of legitimate possible values.

Sometimes, it can be more convenient to preview a dataframe by just looking at the column names. The *colnames()* function is a handy way of doing that.

```{r echo=TRUE}
colnames(winners)
```


### Vectorised Computation

As in a spreadsheet, R dataframes also support *vectorised computation*, in that a particular operation can be applied to each cell in a particular column by applying that function *as if*  we were applying it to just the column.

#### Adding New Columns

As an example of vectorised computation, let's see how to add new columns to a dataframe.

In the first case, consider adding a column that contains the same constant value in each cell. Using a notion known as *broadcast*, we can set the contents of a new column to a single value (a vector of the required length is generated using the value in each location); or we can create vectors of the correct length to add to the dataframe by other means:

```{r}
tmp.df=data.frame(origcol=c('a','b','c'))
tmp.df['newcol']=1
#The nrow() function gives the length (as in, number of rows) of a dataframe
#The seq() function generates a sequence of values of a specified length over a required range
tmp.df$newcol2 = seq(nrow(tmp.df))

tmp.df
```

Alternatively, we can create a new column in the dataframe from the values contained in one or more of the other columns. Let's create a new column, *posdelta*, that contains an integer value that reports on the the number of positions gained, or lost, by each driver going from their starting position on the grid to their final classified position (I'll create a copy of the `winners` dataframe to work with...):

```{r results='asis'}
winners2=winners

winners2['posdelta']=winners['grid']-winners['pos']
#The following is also valid
winners2$posdelta2 = winners2['grid'] - winners2$pos

kable(head( winners2, n=3))
```

The `transform` function in the `plyr` package provides an alternative way of achieving the same thing:

```{r results='asis'}
#Load in the plyr package
library(plyr)

winners3 = transform(winners, posdelta = grid - pos )
kable(head( winners3, n=3))
```

The `transform` function actually allows us to specify multiple columns in the same call:

```{r results='asis'}
winners4 = transform(winners, posdelta = grid - pos, poslapdelta = fastlaprank - pos )
kable(head( winners4, n=3))
```

If you want to create a column that is used to define a further column, use the `mutate()` function rather than `transform`.

Another useful plyr function is `count()` that will count the number of occurrences of unique values of within a dataframe column or set of columns.

### Filtering data in a dataframe

We can filter a dataframe to show just a subset of the data it contains in a variety of ways:

* we can limit the number of rows we want to display
* we can limit the number of columns we want to display
* we can limit the number of rows and the number of columns we want to display

Let's see how to do each of these in turn using two different, but equivalent, techniqes: a "cryptic"" way, and a more verbose way.

We'll start with the verbose way, using the `subset()` function.

```{r results='asis'}
kable( subset(winners, subset=(pos==2 | pos==3)) )
```

The `subset()` function accepts a dataframe, with a *subset* parameter that applies a rule that identifies which rows to include in the returned subset. In this case, I select rows where the *pos* column value equals (==) the number 2 *or* the number 3; (the | symbol is a logical (Boolean) operator that represents the operation OR). We can test for inequlity using the != operator. We can also test for Boolean *and*ed conditions using the operator *&*. Numerical operators can also be applied (for example, *subset=(pos>=2)* would give the same result as shown in the example above).

Cryptically, we can achieve the same thing by passing in a "truth vector" that says whether or not a row is "TRUE" and should be included in the output dataframe:

```{r}
winners[winners['pos']!=1,]
```

Let's unpick that a little. Firstly, we define the "truthiness" of a row condition:

```{r}
winners['pos']!=1
```

Passing this True/False vector into the column slot of the dataframe selects which rows to include in the result (we return the TRUE rows and drop the FALSE ones from the result): *dataframe[rowSelectionVector, columnSelectorList]*. Passing a blank *columnSelectorList* in means return *all* the rows.

To select just a subset of the output columns, pass in a list of the names of the columns we want to select:

```{r}
winners[,c('driverId','race','pos')]
```

Using the subset function, we can identify the columns we want to select using the *select=* parameter.

```{r results='asis'}
kable( subset(winners,select=c(driverId,race,pos)) )
```

We can also select the columns that we *do not* want to include in the output:

```{r results='asis'}
kable( subset(winners,select=-c(driverId,race,pos)) )
```

To both *select* a set of columns and *subset* the rows, we can pass in both arguments:

```{r results='asis'}
kable( subset(winners,subset=(pos==1),select=c(driverId,race)) )
```

Or more cryptically:

```{r results='asis'}
kable( winners[winners['pos']==1,c('driverId','race')] )
```

By carefully choosing the criteria that we use to filter dataframes by row and/or column we can start to have a conversation with a dataframe based on the values contained within it.

### Sorting dataframes

We can also use a variety of sorting operations to sort the data in a dataframe. The `arrange()` function from the `plyr` package provides what is perhaps the easiest way of sorting the rows in a dataframe.

```{r}
#As we have previously loaded the plyr package we should not need to load it again
#library(plyr)

#The arrange function accepts a dataframe and then
#a list of columns to sort on.
arrange(winners[,c('carNum','driverId','race')],carNum)
```

We can sort on multiple columns, in either ascending (*asc*) or descending (*desc*) directions:

```{r}
arrange(winners[,c('carNum','driverId','race')],desc(carNum),race)
```

We can also use the - (minus) operator to specify the reverse, or descending, order direction.

### Merging Dataframes

If we need to merge dataframes, we can do so about a common column. For example, if we split the winners dataframe into two separate, smaller dataframes, we can then merge them back together with the `merge()` function.

```{r}
winners.sub1=winners[,c('driverId','pos')]
winners.sub2=winners[,c('driverId','grid')]
winners.sub.merge=merge(winners.sub1,winners.sub2,by='driverId')
kable(head(winners.sub.merge, n=5))
```

The merge will work even if the columns are ordered differently, and can cope with unmatched entries. If the "merge column" has different names in the two dataframes to be merged, specify the names explicitly using *by.x=* for the name of the merge column in the first dataframe, and *by.y=* for the name of the merge column in the second.

Another form of sorting is to sort the *levels* associated with a factor. The `reorder()` function allows us to reorder the levels of one categorical variable/factor based on the values of a second variable.

### Reshaping Data

Sometimes we may just have a single, long list of data values in an ordered fashion that we want to cast into a tabular format.

In such cases, the following rather handy function will take in the single ordered list of values, along with a list of desired column names, and shape a dataframe for us.

```{r echo=TRUE}
colify=function(datalist,cols){
  df=data.frame(matrix(datalist, nrow=length(datalist)/length(cols), byrow=T))
  names(df)=cols
  df
}

#Example:
#An ordered list of drivers by name, code and manufacturer
ll=c('hamilton','HAM','Mercedes','rosberg','ROS','Mercedes')

#Now let's make a table, specifying the column names we want to apply
colify(ll,c('Name','TLID','Manufacturer'))
```

Several libraries have been developed to help with the reshaping of R dataframes. If you've never really worked with datasets before, it may surprise you to think that a single data set can take on many different shapes. But it can... 

For example, consider this dataframe, which contains the ergast driverId and positions of the podium finishers:

```{r echo=TRUE}
winners.lite=winners[,c('driverId','race','pos')]
head(winners.lite,n=3)
```

This data is in a so-alled *long* form; but there is another equally valid shape that this data can take, often referred to as a *wide* format, in which we have separate columns corresponding to each race, with the values in those columns representing the race winners. To reshape the data, we will use a function from the appropriately named `reshape2` library, which we will need to load in first.

```{r echo=TRUE}
# You may need to download and install the reshape2 package before
# trying to load it in for the first time.

#Once the library is installed, we can load it in
library(reshape2)

wide.df = dcast(winners.lite, driverId ~ race)
head( wide.df, n=3 )
```

In reshaping the data, the `dcast` function identfies the unique values in the the *race* column, and uses these as new column names. It then inspects each row in the original dataframe, looking for unique *driverId* values to act as unique row identifiers in the wide dataframe, then uses (by default) the contents of the rightmost column as the values of the new data cells (in this case, the `pos` column; use `value.var=` to specify another column). If there is no combination of the the row name and new column name in the original dataset, a *not available* NA designator is used.

We can also transform data from *wide* to *long* format, this time using the `melt()` function, which is also contained in the `reshape2` package.

```{r}
#The melt function is also contained within the reshape package.
#However, if we have already loaded the package in, we don't need to load it again
head(melt( wide.df ), n=3)
```

We can also pass in some additional parameters to the `melt()` function that tidy up the resulting data frame

```{r}
melt( wide.df, id.vars=c("driverId"),
      variable.name = "race",
      value.name = "pos",
      na.rm=T )
```

(The *na.rm=T* (or equivalently, *na.rm=TRUE*) parameter setting can be clumsily read as *NA remove is true*, which is to say *remove rows for which it is true that the value is NA*.)


### Split-Apply-Combine

The "split-apply-combine" recipe identified by Hadley Wickham as a widely used technique for extracting groups of a data from a dataset, operating on them, and then combining the results back together, is well supported using his `plyr` R package.

It can be quite a puzzle getting your head round some of the transformations, but there is a logic to it, honestly!

Let's use the `winner` dataframe again, and find the average (mean) position recorded by each team. To do that, we need to *split* the data frame into groups of rows by team, find the average (mean) position for each team (that is, for each separate group of rows) by *applying* the `mean()` function and then create a new dataframe by *combining* rows that contain the names of the teams and their average position. That is, we need to *split - apply - combine*. To do that, we use the `ddply()` function to *summarise* the data:

```{r}
ddply(winners,.(constructorId),summarise, meanpos=mean(pos))
```

The length of the data frame is the same as the number of groups we are summarising.

Another way of using `ddply` is to *mutate* a dataframe. In this case, we can apply a function to each group and use the result in the definition of a new column on the original dataframe (note: the original dataframe is not changed; we need to assign the output of `ddply` to a new variable, or back to the original dataframe):

```{r results='asis'}
winners.new=ddply(winners[,c('carNum','constructorId','pos')],.(constructorId),mutate,meanTeamPos=mean(pos))
winners.new=ddply(winners.new,.(constructorId),mutate,meanTeamPosDelta=pos-mean(pos))
kable(winners.new)
```

`ddply` is a very powerful function for executing *split-apply-combine* style data transformations. If you have ever used spreadsheet pivot tables, you may notice some resemblance to the *summarise* mode of operation. The *mutate* function additionally allows us apply the results of group based summaries to row based operations within each group of rows.

## Summary

This chapter has provided a very quick introduction to the RStudio environment and the R prohramming language.

RStudio is a very powerful tool and can be used to author a wide range of document types that incorporate executable R code and the results of its execution. For example, you can use RStudio to produce slideshows that incorporate R code and the results of running it, or HTML or PDF documents generated from reusable templates. RStudio can also be used to support the development of R packages, which is something I need to learn how to do! To learn more about RStudio, check out the [RStudio documentation](https://support.rstudio.com/hc/en-us/categories/200035113-Documentation), or one of the increasing number of [books on RStudio](http://www.amazon.co.uk/s/ref=nb_sb_noss_1?url=search-alias%3Dstripbooks&field-keywords=rstudio&tag=ouseful-21).

As to the R language itself, we have introduced many of the key tecnhiques and functions that will be used throughout the rest of this book: how to cosntuct dataframes and read data in from CSV files, how to filter dataframes, how to sort (or *arrange*) them, *transform* them, reshape them from wide to long form and back again (*melt* and *dcast*), and process them by group using the *split-apply-combine* method via the `ddply` function.

Over the coming chapters, you will see more of the R language, in particular how it can be used to reshape data to get it into a format where we can easily visualise it. You will also see a lot more of ggplot2, as well as other approaches towards generating data visualisations from within R.

## Exercises

In the `winners` dataframe, what levels are associated with the *constructorId* and *driverId* factors?

Filter the `winners` data frame to generate another dataframe that contains the rows for just the Force India and Mercedes teams. Further limit the derived dataframe to only show the `driverId` and `fastlaptime` columns.
